---
title: "Racial Disparities in Traffic Stops/Citations"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r read-data}
library(tidyverse)
library(infer)
library(broom)
library(Stat2Data)
police <- read.csv("data/nc_durham_2020_04_01.csv")
```

```{r filter-data}
police <- police %>% 
  filter(county_name == "Durham County") %>% 
  filter(subject_race != "unknown" & subject_race != "NA") %>% 
  filter(subject_age != "NA") %>% 
  mutate(subject_race = as.factor(subject_race)) %>% 
  mutate(subject_sex = as.factor(subject_sex))
num_rep <- 10000

```

$~$

## Abstract

Given the unfair nature of the criminal justice system, this paper investigates if there is any evidence of racial bias in traffic stops and citations in Durham County, North Carolina. Specifically, what is the relationship between a subject's demographic attributes (primarily sex and race) and the likelihood of being stopped by police in traffic or receiving a traffic citation in Durham? Our data is an attempted census of individual police stops in Durham created by the Stanford Open Policing project, which collects data on law enforcement nationwide. We hypothesize that race and the likelihood of being stopped by police in traffic or receiving a citation in Durham are associated, with black people disproportionately more stopped relative to their population proportion and more likely to receive a citation upon being stopped. Our conclusions were 1) black people are disproportionately stopped in traffic as compared to their demographic makeup within the Durham population; 2) black people are not in fact the most likely to receive a citation upon being stopped in traffic, rather, Hispanic people are; 3) black females are more likely to be receive a citation upon being stopped than are black males.

\newpage

## Background and Significance

The US incarcerates more people than any other country, and people of color make up a disproportionate percent of the prison population. Police funding has grown significantly over the past four decades, and overpolicing in communities of color is a pressing issue. In Michelle Alexander's book, *The New Jim Crow: Mass Incarceration in the Age of Colorblindness*, she discusses the rise of incarceration rates for black and brown people in the US. Alexander cites the War on Drugs as one of the biggest causes of contemporary mass incarceration, police pretext stops being one example. In pretext stops, cops can pull over a "suspicious" driver on the pretext of a very minor traffic violation (e.g. turning on red, going over the speed limit) and then do a drug sweep of the car, which may result in an arrest for drug-related charges. According to a Pew Research Center survey, "Black adults are about five times as likely as whites to say they’ve been unfairly stopped by police because of their race or ethnicity." We would like to investigate if similar elements of discrimination in traffic stops are evident in Durham County, North Carolina.

There are 326024 observations in our dataset, each an individual police stop recorded in Durham between December 2001 and December 2015. There are `r ncol(police)` variables, relevant variables including `subject_race`, which describes the race of the subject involved in the traffic stop; `subject_sex`, which describes the sex of the subject involved in the traffic stop; `subject_age`, which describes the age of the subject at the time of the traffic stop; `outcome`, which is what resulted from the stop (a warning or a citation, for example); `reason_for_stop`, which describes what the violation leading to the stop was; and `search_conducted`, whether a search of the subject was conducted during the stop. As markers for demographic comparisons, we will be using 2010 Durham census data. Consequently, we filtered our Stanford Open Policing Project dataset to only have values from Durham County, which left us with `r nrow(police)` observations to sample from.


## Methodology

```{r props}
props <- police %>% 
  count(subject_race) %>% 
  mutate(proportion = n/sum(n)) %>% 
  select(subject_race, proportion) %>% 
  mutate(O = "Observed")
```

```{r police-viz}
police_viz <- read.table(text = "
subject_race, proportion, O
asian/pacific islander, 0.056,Expected
black, 0.369,Expected
hispanic, 0.137,Expected
other, 0.009,Expected
white, 0.43,Expected", 
header = TRUE, sep = ",")

# found help from https://stackoverflow.com/questions/16528167/how-to-combine-different-barplot-in-r
# Census data - some overlap with the "Hispanic" ethnicity, as Hispanics can be any race, and the data didn't separate black non-Hispanics from black HIspanics.
# Took out biracial people - not sure where they fitted in
```

```{r combine-data}
combined <- full_join(props, police_viz)
```

```{r props2}
props2 <- police %>% 
  count(subject_sex) %>% 
  mutate(proportion = n/sum(n)) %>% 
  select(subject_sex, proportion) %>% 
  mutate(O = "Observed")
```

```{r police-viz2}
police_viz2 <- read.table(text = "
subject_sex, proportion, O
male, 0.477,Expected
female, 0.523,Expected",
header = TRUE, sep = ",")
```

```{r combine-data2}
combined2 <- full_join(props2, police_viz2)
```


The variables we used to address the research question are `subject_race`, `subject_sex`, `subject_age`, and `citation_issued`. 

To begin, we calculated summary statistics for stop rate based on proportions by race in 2010:

```{r stop-rate}
durham_population <- data.frame(subject_race = c("asian/pacific islander", "black", "hispanic", "other", "white"), population = round(c(0.056 *270001, 0.369* 270001, 0.137*270001, 0.009*270001, 0.43*270001)))

stop_rate <- police %>% 
  filter(year(date) == 2010) %>%
  count(subject_race) %>%
  left_join(durham_population) %>%
  mutate(stop_rate = n / population)

stop_rate %>% 
  kable(digits = 3)
```

From the table, it appears that black people have the highest stop rate among races in Durham County in 2010. This rate is roughly 2.5 times higher than the rate that white people get stopped in Durham County.

To further visualize this trend, we used census data of population proportions by race and compared the expected proportion of people stopped by race to the observed proportion of people stopped by race in Durham County.

$~$

```{r stop-rate-viz, fig.width = 5, fig.height = 3}
ggplot(combined, aes(x = O, y = proportion, fill = subject_race)) + 
  geom_bar(stat = "identity", position = "fill", color = "white") + 
  labs(title = "Proportions of People Stopped by Race", 
       subtitle = "Expected from 2010 Census vs Observed from Data",
       y = "Proportion", 
       x = " ",
       fill = "Subject Race") +
  scale_fill_manual(values = c("#175676", "#4ba3c3", "#cce6f4", "#d62839", "#ba324f"),
                    labels = c("Asian/Pacific Islander", "Black",
                               "Hispanic", "Other", "White")) 
```

$~$

The chart provides some initial evidence for the hypothesis, indicating that a significantly greater proportion of black people were stopped compared to what was expected based on the proportion of black people in Durham County in the 2010 census. Within the "Results" section, we will run statistical tests to determine if the differences between expected and observed proportions were significant.

To begin exploring our second research question on citations, we visualized a segmented bar graph with the proportion of citations based on race below. 

$~$

```{r bargraph, fig.width = 7, fig.height = 3}
ggplot(police, aes(x = subject_race, fill = citation_issued)) + 
  geom_bar(position = "fill") + 
  labs(title = "Likelihood of Citation Based on Race", 
       y = "Proportion", 
       x = "Subject Race", 
       fill = "Citation Issued") + 
  scale_fill_manual(values = c("#ba324f", "#4ba3c3"),
                    labels = c("Not Issued", "Issued")) +
  scale_x_discrete(labels = c("Asian/Pacific Islander",
                              "Black",
                              "Hispanic",
                              "Other",
                              "White")) +
  theme(aspect.ratio = 0.6, axis.text.x = element_text(size = 8)) 
```

$~$

According to the chart, it appears that Hispanics are the race with the highest proportion of citations issued. Surprisingly, black people appear to be the race with the lowest proportion of citations issued. We will further investigate the effect that race plays on the likelihood of receiving a citation by developing a logistic regression model in the second part of our "Results" section.

## Results

### First Research Question (Demographic Factors and Likelihood of Being Stopped)

```{r create-sample}
set.seed(125)
sample <- police %>% 
  group_by(subject_race) %>% 
  sample_frac(size = 0.01) %>% 
  select(subject_race)
# Found code for sampling from https://stackoverflow.com/questions/23479512/stratified-random-sampling-from-data-frame
```

Our exploratory data analysis for the true proportion of black people stopped in traffic indicated that black people appeared to be stopped at a disproportionately higher rate compared to their proportion within the Durham County population. Since the attempted census of `r nrow(police)` observations is far too large to create a bootstrapped null distribution to check the statistical significance of the proportions, we created a stratified proportional sample of `r nrow(sample)` observations, roughly 1% of the original dataset. We decided to check if this difference between the observed and expected proportion (based on Durham County population) was statistically significant through simulation (Appendix 4 for details):

We found that the data provides statistically significant evidence (at the 5% level) that black people are disproportionately more likely to be stopped in Durham County relative to their proportion within the population. The data is also practically significant as indicated by the comparison of stop rates in the exploratory data analysis.

### Second Research Question (Demographic Factors and Likelihood of Citation)

As a preliminary check to decide if further investigation is warranted, we first conducted a chi-squared test for independence, where we found sufficient evidence to indicate the existence of an association between race and citation issued (Appendix 5). We then created a logistic regression model to quantify the effect of race (adjusted for other demographic variables such as sex and age) on log-odds of receiving a citation. The logistic regression model contains four predictors: The subject's race, the subject's age, the subject's sex, and an interaction variable between a subject's race and subject's sex (Appendix 1 for the reasoning behind this logistic regression model, and Appendix 3 for checking the conditions). To determine statistical significance of coefficients, we set the alpha level to 0.01.

```{r mutatestuff}
police1 <- police %>% 
  mutate(subject_race = case_when(
    subject_race == "white" ~ "1white", 
    subject_race == "asian/pacific islander" ~ "asian/pacific islander", 
    subject_race == "black" ~ "black", 
    subject_race == "hispanic" ~ "hispanic", 
    subject_race == "other" ~ "other", 
  ))
```


```{r logreg}
logit_race <- glm(citation_issued ~ subject_race + subject_age + subject_sex +
                    subject_race * subject_sex,
                  data = police1, family = "binomial")
probs <- predict(logit_race, type = "response")
predicted <- ifelse(probs > 0.5, "citation", "no citation")
```



```{r tidy-logit}
tidy(logit_race) %>% 
  kable(digits = 3) 
```

This model yields a few relevant conclusions:

1. Holding age and sex constant, we expect the odds that a Hispanic person will receive a citation upon being stopped by police in Durham County to be `r exp(.357)` times the odds that a white person will receive a citation upon being stopped by police. The coefficient is statistically significant (p-value < 0.01), meaning there is less than a 1% chance such a coefficient or more extreme would be found in the data if race and the likelihood of receiving a citation were not associated. 

2. Holding age and sex constant, we expect the odds that a black person will receive a citation upon being stopped by police in Durham County to be `r exp(-0.127)` times the odds that a white person will receive a citation upon being stopped by police. The coefficient is also statistically significant (p-value < 0.01).

3. In most cases, the interaction variable between sex and race does not result in a statistically significant coefficient, the critical exception being the case of black people. Holding age and race constant, upon being stopped, a black man's odds of receiving a citation are expected to be `r exp(-0.1206676 + 0.032)` times the odds that a black woman will receive a traffic citation upon being stopped. Thus, unlike nearly every other race listed, black people are the only race where women are statistically significantly more likely to receive a citation upon being stopped. 

The implications of this model will be further discussed in the "Discussion" section of the report.

## Discussion

Consistent with our original hypothesis on racial bias in police traffic stops in Durham County, we have learned that black people are disproportionately more likely to be stopped for a traffic violation. This stop rate is both statistically significant and practically relevant, with black people being stopped at a 2.5 times higher rate than white people. Contrary to our hypothesis on racial bias in issuance of citations, however, we have found that black people are disproportionately less likely to receive a citation upon being stopped. Based on our data, we are unable to explain the difference in conclusions between our first and second research question; this may indicate a problem area for further statistical analysis.

Additionally, we found that Hispanics were the most likely to receive a citation upon being stopped. Unfortunately, our observational data cannot create a causal association between a person's race and the likelihood of being stopped or receiving a citation--we only have evidence that suggests statistically significant differences in proportion of black drivers being stopped (as opposed to their population percentage). Nonetheless, our data is consistent with the discussion on racial profiling and pretext stops in Michelle Alexander's *The New Jim Crow: Mass Incarceration in the Age of Colorblindness*, which indicates that black people are more likely to be stopped in traffic and that race does play a role in traffic penalties. There are significant practical implications of this investigation, as racial profiling within the policing system is a huge issue today. Discriminatory incarceration of black people in America severely harms black families, communities, capacity for political participation, etc. Hopefully, larger, more well-established and well-funded initiatives like the Stanford Open Policing Project are able to use similar research to advocate on behalf of criminal justice reform and police reform.

Appendix 2 further explores the trends of the relationship between traffic stops and race over time. Overall, these visualizations show that policing trends by race have not dramatically changed over the time period from 2002 to 2015, which may indicate the enduring relevance of our conclusions today.

For most races, there is very little change in data between citations being given by sex. However, the odds that a black female will receive a citation are significantly greater than their male counterparts. This could possibly show unspoken discrimination against black females, an idea we believe necessitates further statistical analysis. A statistical exploration of stop rates of black females compared to black males and white females could illuminate the unique intersectional experiences faced by black females within America.

A key limitation of our data analysis was that the Stanford Policing Project separated races differently than the census data. Specifically, the Stanford Policing Project data on race did not account for biracial people, people of other races, and did not detail what category white, black, or Asian Hispanics would fall under. As a result, our proportions are slightly skewed. In the future we would attempt to match the categories over the different datasets. If we were able to standardize the groupings we would have less ambiguous data and more accurate proportions and conclusions. Another crucial limitation is that we are unable to establish if the trends we found were causally related. We are only able to take note of trends and cannot isolate causation; thus, our data does not provide conclusive evidence of racial discrimination by police.

\newpage

## References

Alexander, Michelle. *The New Jim Crow: Mass Incarceration in the Age of Colourblindness*. Penguin Books, 2019. 


Barghouty, Pheobe, et al. “The Stanford Open Policing Project.” Openpolicing.stanford.edu, 2020, openpolicing.stanford.edu/. 


“Criminal Justice Facts.” The Sentencing Project, 2 Sept. 2020, www.sentencingproject.org/criminal-justice-facts/. 


DeSilver, Drew, et al. “10 Things We Know about Race and Policing in the U.S.” Pew Research Center, Pew Research Center, 17 Aug. 2020, www.pewresearch.org/fact-tank/2020/06/03/10-things-we-know-about-race-and-policing-in-the-u-s/. 


Rhodes, Nancy, et al. “Risky Driving among Young Male Drivers: The Effects of Mood and Passengers.” Transportation Research Part F: Traffic Psychology and Behaviour, Pergamon, 26 Dec. 2014, www.sciencedirect.com/science/article/pii/S1369847814001727.


“U.S. Census Bureau QuickFacts: Durham County, North Carolina.” Census Bureau QuickFacts, 2010, www.census.gov/quickfacts/fact/table/durhamcountynorthcarolina/RHI225219. 

\newpage

## Appendix 1

This appendix is devoted to showing why we chose the logistic regression model that we did. Utilizing the demographic characteristics of sex, race, and age, we attempted to create the most robust and explanatory model possible from the data. Below are the calculated AIC and BIC values for each model we considered. The model with the lowest AIC and BIC values was the one we chose, as it had the followed the principle of Occam's Razor the most faithfully (it explained the most in the least complex manner).

Logistic regression of Race (BIC, then AIC):

```{r aic1}
logit_raceonly <- glm(citation_issued ~ subject_race, data = police, 
                      family = "binomial")
BIC(logit_raceonly)
AIC(logit_raceonly)
```

Logistic regression of Race and Age:

```{r aic2}
logit_race_age <- glm(citation_issued ~ subject_race + subject_age,
                      data = police, family = "binomial")
BIC(logit_race_age)
AIC(logit_race_age)
```

Logistic Regression of Race, Sex, and Age: 

```{r aic3}
logit_race_sex <- glm(citation_issued ~ subject_race + subject_sex + 
                        subject_age, data = police, family = "binomial")
BIC(logit_race_sex)
AIC(logit_race_sex)
```

Noting that the best logistic regression included all three variables (despite the fact that sex by itself is not significantly associated with the likelihood of receiving a citation), we tested interaction variables.

Logistic Regression of Race, Sex, Age, and Race * Age:

```{r aic4}
logit_race_times_age <- glm(citation_issued ~ subject_race + subject_sex + 
                        subject_age + subject_race * subject_age,
                        data = police, family = "binomial")
BIC(logit_race_times_age)
AIC(logit_race_times_age)
```

BIC value increased, so we eliminated the interaction variable.

Logistic Regression of Race, Sex, Age, and Age * Sex:

```{r aic5}
logit_age_times_sex <- glm(citation_issued ~ subject_race + subject_sex + 
                        subject_age + subject_age * subject_sex,
                        data = police, family = "binomial")
BIC(logit_age_times_sex)
AIC(logit_age_times_sex)
```

BIC again increased, so we eliminated the interaction variable.

Logistic Regression of Race, Sex, Age, and Race * Sex:

```{r aic6}
BIC(logit_race)
AIC(logit_race)
```

We obtained both our lowest AIC value here and a lower BIC value, making this our most robust and explanatory logistic regression model.

## Appendix 2

```{r line-graph1}
library(lubridate)
line <- police %>% 
  filter(subject_race != "other", year(date) != 2001) %>% 
  mutate(year = year(date))

line <- line %>% 
  group_by(year) %>% 
  count(subject_race) %>% 
  mutate(prop = n/sum(n)) 

```

```{r line-graph2, fig.height = 4, fig.width = 6}
ggplot(line, aes(x = year, y = prop, color = subject_race)) +
  geom_point() + 
  geom_line() + 
  labs(title = "Proportions Stopped by Race over Time", 
       x = "Year",
       y = "Proportion Stopped in Traffic", 
       color = "Subject Race") + 
  scale_color_manual(values = c("#175676", "#d62839", "#4ba3c3", "#ba324f")) +
  theme(aspect.ratio = 0.8)
```

## Appendix 3

Conditions of Logistic Regression:

1. Independence - Each traffic stop is independent of other traffic stops; one traffic stop resulting in a citation does not affect the likelihood that other traffic stops result in citations.

2. Linearity - Below, we have depicted scatterplots of the relationship between age and the log-odds of receiving a citation. The Linearity Assumption is met because there is a roughly linear relationship between the age of a subject (the quantitative predictor) and the log-odds of receiving a citation.

```{r emplogits, fig.width = 4, fig.height = 4}
emplogitplot1(citation_issued ~ subject_age, 
              data = police, ngroups = 10,
              main = "Empirical Logits vs Age")
```

3. Randomness - We do not have reason to believe that the attempted census of police stops would have results that differ substantially with a full census of police stops.

Conditions met. Proceed with a logistic regression model. 

## Appendix 4: Simulation-Based Hypothesis Test

Let $\rho$ equal the true proportion of stopped drivers who were black within Durham County. 

$H_0: \rho = 0.369$. The true proportion of stopped drivers who were black within Durham County is equal to the true proportion of black people within Durham County (0.369).

$H_A: \rho > 0.369$. The true proportion of stopped drivers who were black within Durham County is greater than the true proportion of black people within Durham County.

$\alpha$ = 0.05

```{r mutate-count}
sample <- sample %>% 
  mutate(B = case_when(
    subject_race == "black" ~ "B",
    subject_race != "black" ~ "Not B"
  ))
```

```{r simulate-null}
set.seed(400)
null_dist1 <- sample %>% 
  specify(response = B, success = "B") %>% 
  hypothesize(null = "point", 
              p = c("B" = 0.369, "Not B" = 0.631)) %>% 
  generate(reps = num_rep, type = "simulate") %>% 
  calculate(stat = "prop")
```
```{r obs-prop}
obs_prop <- sample %>% 
  specify(response = B, success = "B") %>% 
  calculate(stat = "prop") %>% 
  pull()
```

```{r calc-p-value}
p <- null_dist1 %>% 
  filter(stat >= obs_prop) %>% 
  summarise(p = n()/nrow(null_dist1)) %>% 
  pull()
p
```

Because our p-value of `r round(p, 3)` is less than our $\alpha$ of 0.05, we reject the null hypothesis. There is sufficient evidence to indicate that the true proportion of people who are stopped within Durham County that are black is greater than the proportion of black people within the Durham County population (0.369). This indicates that black people are disproportionately stopped at a higher rate.

## Appendix 5: Chi-Squared Test

We conducted a chi-squared test of independence to determine if a person's race is associated with a higher chance of receiving a citation upon being stopped.

$H_0:$ Race and the likelihood of receiving a citation upon being stopped are not associated.

$H_A:$ Race and the likelihood of receiving a citation upon being stopped are associated.

$\alpha$ = 0.05.

```{r chisq}
chi1 <- police %>% 
  chisq_test(formula = citation_issued ~ subject_race)
chi1
```

The chi-squared test for independence outputted a statistic of `r round(chi1$statistic, 3)`. The distribution of the test statistic is a chi-squared distribution, which is unimodal and right-skewed with 4 degrees of freedom.

#### Conclusion:
Since our p-value of `r round(chi1$p_value, 3)` is less than our $\alpha$ of 0.05, we reject the null hypothesis. There is sufficient evidence to indicate that race and the likelihood of receiving a citation upon being stopped are associated.


